Covariant Phase Space via MultiSymplectic Geometry

### 2017-05-25 cpsviams.tex
Almost finished....

\usepackage{amsmath}
\usepackage[multisym, geomec,basic, diffgeo]{./Math-Symbols-List/toninus-math-symbols}
\usepackage[subpreambles=true]{standalone}
\usepackage{graphicx}
\usepackage{wrapfig}

@@

\begin{abstract}
This is the paper's abstract \ldots
\end{abstract}

@@

\item It turns out that  $\LinDualJet $ carries a naturally defined multisymplectic form $\omega = -\textrm{d} \theta$ derived by a multicanonical m-form $\theta$. \\
	(This has been obtained pulling back the tautological m-form $\theta_\Lambda$ from $\bigwedge^m (E)$.) \\
	Thus, it can be considered the theoretical analogue of the cotangent bundle in ordinary geometric mechanics.

@@

\item Via $\mathbb{F} \Lagrangian$ we can pull-back the canonical forms on $J^1 E$ to give the \emph{Poincarè - Cartan forms}
	\begin{displaymath}
		\theta_\Lagrangian = \mathbb{F} \Lagrangian^\ast \theta \qquad \omega_\Lagrangian = - \textrm{d} \theta_\Lagrangian
	\end{displaymath}
	and via $\Hamiltonian$ we can pull-back the canonical forms on $\LinDualJet$ to give the \emph{De Donder - Weyl forms}
	\begin{displaymath}
		\theta_\Hamiltonian = \Hamiltonian^\ast \theta \qquad \omega_\Hamiltonian = - \textrm{d} \theta_\Lagrangian
	\end{displaymath}

@@

	\begin{definition}[Jacobi operator pertaining to $D_\cdot$ around a solution $\phi \in \Sol$]\label{Def:JacobiOp}
		\begin{displaymath}
			J(\phi) \; : \; T_\phi \Conf \rightarrow	\phi^\ast(V^\TwistedLinear F) 
			\quad s.t. \quad  \left(J(\phi)\right) [ \delta\phi ]
			= \sigma \left( \left.\dfrac{\textrm{d}}{\textrm{d} \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0} \right)
		\end{displaymath}
	\end{definition}

\begin{figure}
\centering
\includegraphics[width=0.40\textwidth]{Pictures/Jacobi.png}
\caption{\footnotesize \label{fig:jacobi}$ \left(J(\phi)\right) [ \delta\phi ]$ and $D$ target on the same space.}
\end{figure}
where:
\begin{itemize}
	\item $\phi_\lambda$ is an arbitrary variation of $\phi_0 = \phi$ generated by $\delta\phi$.
	\item exactly as noted for the tangent vector of $\Conf$, we have for any $x\in M$:
		\begin{displaymath}
			\left.\dfrac{\textrm{d}}{\textrm{d} \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0}(x) \in V_M W
			\quad \Rightarrow \quad
			\left.\dfrac{\textrm{d}}{\textrm{d} \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0} \in D[\phi]^\ast \left( V_M \left( V^\TwistedLinear \right)\right)
		\end{displaymath}
		because $D_\cdot[\phi_\lambda]$ can be regarded as a section of $V^\TwistedLinear F$ over $M$, 
		thus the latter definition does not depend from the particular choice of variation $\phi_\lambda$ 
		but only from its infinitesimal generator $\delta\phi$.
	\item $\sigma$ is the canonical bundle-morphism
		\begin{displaymath}
			\sigma:  0^\ast \left( V_M W \right)  \; \rightarrow \;  0^\ast \left(V_E W \right) \simeq W
		\end{displaymath}
		 where $V_M W$ and $V_F W$ are the vertical sub-bundle of $T W$ w.r.t. $M$ and $F$ respectively 
		 and $0$ is the zero-section of $W$ over $F$.
\end{itemize}

@@

\begin{proposition}\label{Prop:CurrentClosed}
	When restricted to $\textrm{T}_\phi \Sol$, the symplectic current is a closed form. That is:
	\begin{displaymath}
		\textrm{d} \Upsilon_\phi \left( \delta \phi_1 , \delta \phi_2 \right) = 0 \qquad \delta \phi_1 , \delta \phi_2 \in T_\phi \Sol
	\end{displaymath}
\end{proposition}
\begin{proof}
	Working in the collective notation introduced above, we have:
	\begin{align*}
		\textrm{d} \Upsilon_\phi \left( \delta \phi_1 , \delta \phi_2 \right) &= 
		\phi^\ast \textrm{d} \left[ i_{\delta\phi_2} i_{\delta \phi_1} \omega \right] = 
		\phi^\ast \left( - i_{\delta\phi_2} \textrm{d} ( i_{\delta \phi_1} \omega )  + \Lie_{\delta\phi_2 } ( i_{\delta \phi_1} \omega ) \right)
	\end{align*}

\end{proof}

@@

\begin{proposition}
	In this condition $\Omega_\Sigma (\phi): T_\phi \Sol \times T_\phi \Sol \rightarrow \Real$ does not depends on the choice of $\Sigma$ anymore.
\end{proposition}
\begin{proof}
	Consider 
	\begin{displaymath}
		K_t = \supp\left(\Upsilon_\phi (\delta \phi_1, \delta \phi_2) \right) \cap \Gamma_t
	\end{displaymath}
	by varying $t$, it determines an m-dimensional domain $V \subset M$.
	For any $U$ open set in $M$ containing $K={K_t}_{0<t<t'}$ we have, as a simple consequence of proposition \ref{Prop:CurrentClosed} and Stokes' lemma, that
	\begin{displaymath}
		\int_V \textrm{d} \Upsilon_\phi = - \int_{\Gamma_0} \Upsilon_\phi +\int_{\Gamma_t} \Upsilon_\phi 
	\end{displaymath}
	Therefore $\Omega_\Sigma (\phi) = \Omega_{\Sigma'} (\phi)$ for all possible pair of Cauchy surfaces.
\end{proof}

>>>>

\usepackage{amsmath}
\usepackage[multisym, geomec, basic, diffgeo]{./Math-Symbols-List/toninus-math-symbols}
\usepackage[subpreambles=true]{standalone}
\usepackage{commath}

\usepackage{graphicx}
\usepackage{wrapfig}

@@

\begin{abstract}
La costruzione del covariant phase space con la cognizione del formalismo simplettico come descritta nell'articolo di forger romero.

Piccola rivisisazione sulla dimostrazione della corrente simplettica (indipendentemente dal formalismo adottato)

Con lo stesso linguaggio si può descrivere più precisamente l'algoritmo di Peierls partendo dalle variazione di soluzioni sottomesse ad una variazione di lagrangiana.

\end{abstract}

@@

\item It turns out that  $\LinDualJet $ carries a naturally defined multisymplectic form $\omega = -\ExtD \theta$ derived by a multicanonical m-form $\theta$. \\
	(This has been obtained pulling back the tautological m-form $\theta_\Lambda$ from $\bigwedge^m (E)$.) \\
	Thus, it can be considered the theoretical analogue of the cotangent bundle in ordinary geometric mechanics.

@@

\item Via $\mathbb{F} \Lagrangian$ we can pull-back the canonical forms on $J^1 E$ to give the \emph{Poincarè - Cartan forms}
	\begin{displaymath}
		\theta_\Lagrangian = \mathbb{F} \Lagrangian^\ast \theta \qquad \omega_\Lagrangian = - \diff \theta_\Lagrangian
	\end{displaymath}
	and via $\Hamiltonian$ we can pull-back the canonical forms on $\LinDualJet$ to give the \emph{De Donder - Weyl forms}
	\begin{displaymath}
		\theta_\Hamiltonian = \Hamiltonian^\ast \theta \qquad \omega_\Hamiltonian = - \diff \theta_\Lagrangian
	\end{displaymath}

@@

	\begin{definition}[Jacobi operator pertaining to $D_\cdot$ around a solution $\phi \in \Sol$]\label{Def:JacobiOp}
		\begin{displaymath}
			J(\phi) \; : \; T_\phi \Conf \rightarrow	\phi^\ast(V^\TwistedLinear F) 
			\quad s.t. \quad  \left(J(\phi)\right) [ \delta\phi ]
			= \sigma \left( \left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0} \right)
		\end{displaymath}
	\end{definition}

\begin{figure}
\centering
\includegraphics[width=0.40\textwidth]{Pictures/Jacobi.png}
\caption{\footnotesize \label{fig:jacobi}$ \left(J(\phi)\right) [ \delta\phi ]$ and $D$ target on the same space.}
\end{figure}
where:
\begin{itemize}
	\item $\phi_\lambda$ is an arbitrary variation of $\phi_0 = \phi$ generated by $\delta\phi$.
	\item exactly as noted for the tangent vector of $\Conf$, we have for any $x\in M$:
		\begin{displaymath}
			\left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0}(x) \in V_M W
			\quad \Rightarrow \quad
			\left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0} \in D[\phi]^\ast \left( V_M \left( V^\TwistedLinear \right)\right)
		\end{displaymath}
		because $D_\cdot[\phi_\lambda]$ can be regarded as a section of $V^\TwistedLinear F$ over $M$, 
		thus the latter definition does not depend from the particular choice of variation $\phi_\lambda$ 
		but only from its infinitesimal generator $\delta\phi$.
	\item $\sigma$ is the canonical bundle-morphism
		\begin{displaymath}
			\sigma:  0^\ast \left( V_M W \right)  \; \rightarrow \;  0^\ast \left(V_E W \right) \simeq W
		\end{displaymath}
		 where $V_M W$ and $V_F W$ are the vertical sub-bundle of $T W$ w.r.t. $M$ and $F$ respectively 
		 and $0$ is the zero-section of $W$ over $F$.
\end{itemize}

@@

\begin{proposition}\label{Prop:CurrentClosed}
	When restricted to $\textrm{T}_\phi \Sol$, the symplectic current is a closed form. That is:
	\begin{displaymath}
		\diff \Upsilon_\phi \left( \delta \phi_1 , \delta \phi_2 \right) = 0 \qquad \delta \phi_1 , \delta \phi_2 \in T_\phi \Sol
	\end{displaymath}
\end{proposition}
\begin{proof}
	Working in the collective notation introduced above, considering $\phi \in \Sol$ and $V_i \in T_\phi \Sol$  we have:
	\begin{align*}
		\diff \Upsilon_\phi \left( V_1 , V_2 \right) &= 
		\phi^\ast \diff \left[ i_{V_2} i_{V_1} \omega \right] = 
		\phi^\ast \left( - i_{V_2} \diff ( i_{V_1} \omega )  + \Lie_{V_2 } ( i_{V_1} \omega ) \right) =	\\
		&= \phi^\ast \left( 
			i_{V_2}i_{V_1} \diff \omega - 
			\Lie_{V_1 } ( i_{V_2} \omega ) +
			\Lie_{V_2 } ( i_{V_1} \omega ) +
			i_{[V_1,V_2]} \omega
		\right)
	\end{align*}
	using the Cartan's magic formula twice and the commutation rule between Lie derivative and contraction operator $i_\cdot$.\\
	The first term vanishes because $\omega= \diff \theta$ is exact by construction.\\
	The second and the third term vanish because
	\begin{displaymath}
		\left\langle \left( J(\phi) \right)[V_i] \right\vert \left. V_j \right\rangle = 
		\phi^\ast \Lie_{V_i} \left( i_{V_j} \omega \right)
	\end{displaymath}
	and $V_i$ are solution of the Jacobi equation
\end{proof}

@@

\begin{proposition}
	In this condition $\Omega_\Sigma (\phi): T_\phi \Sol \times T_\phi \Sol \rightarrow \Real$ does not depends on the choice of $\Sigma$ anymore.
\end{proposition}
\begin{proof}
	Consider 
	\begin{displaymath}
		K_t = \supp\left(\Upsilon_\phi (\delta \phi_1, \delta \phi_2) \right) \cap \Gamma_t
	\end{displaymath}
	by varying $t$, it determines an m-dimensional domain $V \subset M$.
	For any $U$ open set in $M$ containing $K={K_t}_{0<t<t'}$ we have, as a simple consequence of proposition \ref{Prop:CurrentClosed} and Stokes' lemma, that
	\begin{displaymath}
		\int_V \diff \Upsilon_\phi = - \int_{\Gamma_0} \Upsilon_\phi +\int_{\Gamma_t} \Upsilon_\phi 
	\end{displaymath}
	Therefore $\Omega_\Sigma (\phi) = \Omega_{\Sigma'} (\phi)$ for all possible pair of Cauchy surfaces.
\end{proof}

\subsection{Example: Hamiltonian Picture}
Let's work out all the calculation for the specific case: Hamiltonian dynamics.\\
Consider the following trivializing coordinates on the multiphase space:
\begin{displaymath}
	\LinDualJet \rightsquigarrow (x^\mu, q^A, p^\mu_A)
\end{displaymath}
%
Dynamics is encoded into the Hamiltonian density $H$:
\begin{displaymath}
	\mathcal{H}(x^\mu, q^A, p^\mu_A) = 
	\left(x^\mu, q^A, p^\mu_A, - H( x^\mu, q^A, p^\mu_A) \: \diff^m x \right)
\end{displaymath}
%
The De Donder - Weyl forms are obtained by the tautological form on $\AffDualJet \rightsquigarrow (x^\mu, q^A, p^\mu_A, p)$
\begin{displaymath}
	\theta = p^\mu_A \diff q^A \wedge \diff^m x_\mu + p \diff^m x
\end{displaymath}
via pull-back
\begin{displaymath}
	\theta_\Hamiltonian = p^\mu_A \diff q^a \wedge \diff^m x_\mu - H \diff^m x
\end{displaymath}
and external derivation
\begin{displaymath}
	\omega_\Hamiltonian = -\diff \theta_\Hamiltonian = \diff q^A \wedge \diff p^\mu_A \wedge \diff^m x_\mu + \diff H \wedge \diff^m x 
\end{displaymath}
%
Contracting $\omega_\Hamiltonian$ with a vertical (w.r.t. $M$) tangent field on $\LinDualJet$
\begin{displaymath}
	V = V^A \dfrac{\partial}{\partial q^A} +
			V^\mu_A \dfrac{\partial}{\partial p^\mu_A}
\end{displaymath}
and then pulling back with a section $(\varphi,\pi)$ gives:
\begin{align*}
	(\varphi , \pi)^\ast (i_V \omega_\Hamiltonian ) = & 
	\partial_\mu \pi^\mu_A \, V^A(\varphi,\pi) \, \diff^m x 
	+ \frac{\partial H}{\partial q^A} (\varphi,\pi) \, V^A(\varphi,\pi) \, \diff^m x + \\
	 & - \partial_\mu \varphi^A \, V^\mu_A(\varphi,\pi) \, \diff^m x +
	\frac{\partial H}{\partial p^\mu_A} (\varphi,\pi) \, V^\mu_A(\varphi,\pi) \, \diff^m x
\end{align*}
where $(\varphi, \pi)$ has to be intended as $\left(x^\mu, \varphi^A(x), \pi^\mu_A(x)\right)$.
%
This leads to an explicit formula for the De Donder Weyl map:
\begin{align*}
	\eval{\mathcal{D}_\Hamiltonian \left( j^1 (\varphi, \pi) \right) }_{x} = \biggr( & 
		x^\mu, \varphi^A(x), \pi^\mu_A(x), \\
		& \left( \frac{\partial H}{\partial q^A} (\varphi,\pi) + \partial_\mu \pi^\mu_A \right)	\diff q^A \otimes \diff^m x 
			+ \left( \frac{\partial H}{\partial p^\mu_A} (\varphi,\pi) - \partial_\mu \varphi^A \right)	\diff p^\mu_A \otimes \diff^m x
		\biggr)
\end{align*}
%
Considering then a variation of sections $(\varphi_\lambda, \pi_\lambda)$ and denoting
\begin{displaymath}
	\delta \varphi = 
	\eval{ \dfrac{\partial}{\partial \lambda} \varphi_\lambda }_{ \lambda=0} =
	\delta \varphi^i \partial_i	\\
	\; , \qquad
	\delta \pi = 
	\eval{ \dfrac{\partial}{\partial \lambda} \pi_\lambda }_{ \lambda=0} =
	\delta \pi^i \partial_i
\end{displaymath}
can be explicitly compute the directional derivative of the map
\begin{align*}
	\eval{ \dfrac{\partial}{\partial \lambda} \mathcal{D}_\Hamiltonian \left( j^1 (\varphi, \pi) \right)}_{ \lambda=0} = &
	\delta \varphi^A \, \dfrac{\partial}{\partial q^A} + 
	\delta \pi^\mu_A \dfrac{\partial}{\partial p^\mu_A} + \\
	& + \left( 
		\frac{\partial^2 H}{\partial q^B \partial q^A} (\varphi, \pi) \delta \varphi^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial q^A} (\varphi, \pi) \delta \pi^\nu_B +
		\partial_\mu \delta \pi^\mu_A
	\right)\diff q^A  \otimes \diff^m x + \\
	& + \left( 
		\frac{\partial^2 H}{\partial q^B \partial p^\mu_A} (\varphi, \pi) \delta \varphi^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial p^\mu_A}(\varphi, \pi) \delta \pi^\nu_B -
		\partial_\mu \delta \varphi^A
	\right)\diff p^\mu_A  \otimes \diff^m x
\end{align*}
where the last two terms determine an explicit expression for the Jacobi operator in the case that $(\varphi,\pi)$ is a solution of the motion equations:
\begin{align*}
	J_\Hamiltonian\left( (\varphi, \pi)\right) \left[ (\delta\varphi , \delta \pi ) \right] =
	& \left( 
		\frac{\partial^2 H}{\partial q^B \partial q^A} (\varphi, \pi) \delta \varphi^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial q^A} (\varphi, \pi) \delta \pi^\nu_B +
		\partial_\mu \delta \pi^\mu_A
	\right)\diff q^A  \otimes \diff^m x + \\
	& + \left( 
		\frac{\partial^2 H}{\partial q^B \partial p^\mu_A} (\varphi, \pi) \delta \varphi^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial p^\mu_A}(\varphi, \pi) \delta \pi^\nu_B -
		\partial_\mu \delta \varphi^A
	\right)\diff p^\mu_A  \otimes \diff^m x
\end{align*}
thus a variation $(\delta \varphi, \delta \pi)$ lies in $\Sol$ if
\begin{align}\label{Eq.:Jacobi}
	\partial_\mu \delta \pi^\mu_A &=
	- \frac{\partial^2 H}{\partial q^B \partial q^A} (\varphi, \pi) \delta \varphi^B
	- \frac{\partial^2 H}{\partial p^\nu_B \partial q^A} (\varphi, \pi) \delta \pi^\nu_B \\
		\partial_\mu \delta \varphi^A &=
		\frac{\partial^2 H}{\partial q^B \partial p^\mu_A} (\varphi, \pi) \delta \varphi^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial p^\mu_A}(\varphi, \pi) \delta \pi^\nu_B
\end{align}
%
Finally, considering two variation $(\delta \varphi_i, \delta \pi_i)$ around a solution $(\varphi,\pi)$ we can compute the symplectic current:
\begin{align*}
	\Upsilon_{(\varphi,\pi)} \left( (\delta \varphi_1, \delta \pi_1) , (\delta \varphi_2, \delta \pi_2) \right) = &
	(\varphi, \pi)^\ast \left[ i_{( \delta \varphi_2, \delta \pi_2)} i_{( \delta \varphi_1, \delta \pi_1)} \omega_\Hamiltonian \right] = \\
	= & (\varphi, \pi)^\ast	\left[ (\delta \varphi_1)^A (\delta \pi_2)^\mu_A - (\delta \varphi_2)^A (\delta \pi_1)^\mu_A \right]	\diff^m x_\mu = \\
	= & \Upsilon_{(\varphi,\pi)}^\mu \diff \sigma_\mu
\end{align*}
where $\diff \sigma_\mu $ is the induced volume form on $\Sigma$ from $M$.\\
Furthermore the external derivative of the form results:
\begin{align*}
	\diff \Upsilon_{(\varphi,\pi)} &\left( (\delta \varphi_1, \delta \pi_1) , (\delta \varphi_2, \delta \pi_2) \right)  =
	\partial_\mu \Upsilon_{(\varphi,\pi)}^\mu \diff^mx = \\
	& = (\partial_\mu \delta \varphi_1)^A (\delta \pi_2)^\mu_A - (\partial_\mu \delta \varphi_2)^A (\delta \pi_1)^\mu_A + 
	(\delta \varphi_1)^A (\partial_\mu  \delta \pi_2)^\mu_A - (\delta \varphi_2)^A (\partial_\mu  \delta \pi_1)^\mu_A
\end{align*}
In the case that $(\delta \varphi_i, \delta \pi_i) \in T_{(\varphi,\pi)}\Sol$ we can plug equations \ref{Eq.:Jacobi} in the latter expression giving
\begin{align*}
	\partial_\mu \Upsilon_{(\varphi,\pi)}^\mu  =&	 
	\left(
		\frac{\partial^2 H}{\partial q^B \partial p^\mu_A} (\varphi,\pi) \delta \varphi_1^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial p^\mu_A} (\varphi,\pi) ( \delta \pi_1)^\nu_B +
	\right) ( \delta \pi_2)^\mu_A +\\
	&-\left(
		\frac{\partial^2 H}{\partial q^B \partial q^A} (\varphi,\pi) \delta \varphi_2^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial q^A} (\varphi,\pi) ( \delta \pi_2)^\nu_B +
	\right) ( \delta \varphi_1)^A +\\
	&-\left(
		\frac{\partial^2 H}{\partial q^B \partial p^\mu_A} (\varphi,\pi) \delta \varphi_2^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial p^\mu_A} (\varphi,\pi) ( \delta \pi_2)^\nu_B +
	\right) ( \delta \pi_1)^\mu_A +\\
	&+\left(
		\frac{\partial^2 H}{\partial q^B \partial q^A} (\varphi,\pi) \delta \varphi_1^B +
		\frac{\partial^2 H}{\partial p^\nu_B \partial q^A} (\varphi,\pi) ( \delta \pi_1)^\nu_B
	\right) ( \delta \varphi_2)^A \\
	&=0
\end{align*}
thus proving the closedness of the symplectic current.
